Full Title

### 2018-01-24 cones/paper/cones.tex
Made changes.

\section{PCF with Countable Nondeterminism}

The language we shall study is similar to that given in \cite{LairdOrdinalGames}.  
It is the language formed by adding an unbounded natural-number choice primitive on to the base language PCF.  
That is, it is the language formed by taking the simply-typed $\lambda$-calculus over the ground type $\nat$, and adding the constants:
\begin{itemize}
  \item Structural constants $\0\from\nat$, $\suc,\pred\from\nat\to\nat$ and $\IfO \from \nat \to \nat \to \nat \to \nat$ for $\nat$
  \item Recursion combinators $\Y_T \from (T \to T) \to T$ for each $T$.
  \item Countable choice $\wn\from\nat$.
\end{itemize}

A small-step operational semantics for this language, taken from \cite{LairdOrdinalGames}, is given in Figure \ref{fig:op-sem}.
\begin{figure}
  \begin{mathpar}
    \inferrule*{ }{(\lambda x.M) N \opto M[N/x]}
    \and
    \inferrule*[right=$n\in\bN$]{ }{\wn\opto\n}
    \and
    \inferrule*{ }{\Y M \opto M (\Y M)}
    \and
    \inferrule*{ }{\IfO \0\;x\;y \opto x}
    \and
    \inferrule*{ }{\IfO (\suc \n)\;x\;y \opto y}
    \and
    \inferrule*{M \opto M'}{M N \opto M' N}
    \and
    \inferrule*{M \opto M'}{\suc M \opto \suc M'}
    \and
    \inferrule*{M \opto M'}{\pred M \opto \pred M'}
    \and
    \inferrule*{M \opto M'}{\IfO M \opto \IfO M'}
  \end{mathpar}
  \caption{Operational Semantics for a PCF-like language with countable nondeterminism.}
  \label{fig:op-sem}
\end{figure}
Our main task will be to give a Denotational Semantics for this language.

Given a term $M$ of ground type $\nat$, we say that $M$ \emph{must converge}, and write $M\mustconverge$, if every evaluation $\pi$ of $M$ (according to the operational semantics given in Figure \ref{fig:op-sem}) terminates at some numeral value.  
We write $M \converges \n$ if there exists some evaluation $\pi$ of $M$ that terminates at $\n$.

Given a type $T$, we define an order on the terms of type $T$, by saying that $M\lmam N$ if for all compatible contexts $C[-]$ of type $\nat$ we have
\begin{gather*}
  C[M]\converges\n\Rightarrow C[N]\converges \n \\
  C[M]\mustconverge\Rightarrow C[N]\mustconverge\,.
\end{gather*}

We say $M\emam N$ if $M\lmam N$ and $N\lmam M$.

@@

\section{The base model}

We will give a model based on Abramsky, Jagadeesan and Malacaria's Game Semantics for PCF \cite{ajmPcf}.  
We do this partly for the sake of simplicity, and partly to demonstrate the power of our methods -- since AJM games are typically seen as less flexible than the Hyland-Ong games of \cite{hoPcf}.  
However, we stress that our construction can be carried out just as well in Hyland and Ong's model, and we will keep this model in mind throughout, and mention whenever a slightly modified argument is needed for that model.

\subsection{Games}

A \emph{game} is a tuple $A=(M_A,\lambda_A,P_A)$, where
\begin{itemize}
  \item $M_A$ is a set of \emph{moves},
  \item $\lambda_A\from M_A \to \OP\times\QA$ is a function that designates each move as either a \emph{$P$-move} or an \emph{$O$-move}, and as either a \emph{question} or an \emph{answer}.
  \item $P_A$ is a non-empty prefix-closed set of finite sequences of elements of $M_A$ called \emph{positions}, satisfying the following two conditions:
    \begin{description}
      \item[Alternation] any non-empty position in $P_A$ starts with an $O$-move, and if $sab\in P_A$ then $a$ and $b$ have opposite signs - one is a $P$-move and the other an $O$-move.  
      \item[Well-bracketing] If $s\in P_A$, then the corresponding $QA$-sequence of questions and answers is well-bracketed.
    \end{description}
\end{itemize}

\subsection{Strategies}

A (deterministic) \emph{strategy} for a game $A$ is a non-empty, prefix-closed subset $sigma\subset P_A^+$ satisfying \emph{determinism}: if $sa,sb\in \sigma$ are both $P$-positions, then $a=b$.  
Here, $P_A^+$ is the set of $P$-positions in $A$.

We say that a strategy $\sigma$ is \emph{history-free} if whenever $sab,tac\in\sigma$ are both $P$-positions, then $b=c$.  
Such a strategy can be though of as a partial function $M_A^-\pfun M_A^+$, where $M_A^-$ is the set of $O$-moves and $M_A^+$ the set of $P$-moves in $A$.  
If this function can be given as a partial recursive function, we say that the strategy $\sigma$ is \emph{recursive}.

\subsection{Connectives}

Given games $A$ and $B$, we can define games $A\times B$, $A\tensor B$ and $A\implies B$ as follows:
\begin{itemize}
  \item $M_{A\times B}=M_{A\tensor B}=M_{A\implies B}=M_A+M_B$.  
  \item $\lambda_{A\times B}=\lambda_{A\tensor B}=[\lambda_A,\lambda_B]$.  
    $\lambda_{A\implies B}=[\neg\circ\lambda_A,\lambda_B]$, where $\neg\from\OP\times\QA\to\OP\times\QA$ is the function that exchanges $O$ and $P$ and leaves $Q$ and $A$ alone.
  \item $P_{A\times B}$ is the union of $P_A$ and $P_B$.  
    $P_{A\tensor B}$ and $P_{A\implies B}$ are defined by the following formula:
    \[
      \{s\in (M_A + M_B)^* \suchthat s\vert_A\in P_A,\;s\vert_B\in P_B,\;\text{$s$ is alternating}\}\,.
    \]
  Note that this is a different strategy for $P_{A\tensor B}$ and $P_{A\implies B}$, because the definition of an `alternating sequence' depends on the sign function $\lambda$.
\end{itemize}

\subsection{Composition of strategies}

Given strategies $\sigma\from A\implies B$ and $\tau\from B\implies C$, we define a set
\[
  \sigma\|\tau = \{\s\in(M_A+M_B+M_C)^*\suchthat \s\vert_{A,B}\in\sigma,\;\s\vert_{B,C}\in\tau\}\,.
  \]
We then define
\[
  \sigma;\tau = \{\s\vert_{A,C}\suchthat \s\in\sigma\|\tau\}\,.
  \]
It may be proved \cite{ajmPcf} that this notion of composition is associative and has identities given by the obvious copycat strategy for the game $A\implies A$.  
We therefore end up with a category whose objects are games and where the morphisms from a game $A$ to a game $B$ are the strategies for the game $A\implies B$.

It can also be shown that the composition of two history-free strategies is history free, and that the composition of two recursive strategies is recursive.  
Furthermore, the identity strategy is always history-free and recursive.

We write $\G$ for the category of games and strategies, $\G_{hf}$ for the category of games and history-free strategies, and $\G_{rec}$ for the category of games and recursive strategies.

Lastly, it is well-known that the connectives $\tensor$ and $\implies$ exhibit $\G$ as a monoidal closed category (with tensor unit the empty game $I$ with no moves) and that $\times$ gives products in $\G$.  
All these properties pass on to $\G_{hf}$ and $\G_{rec}$ in the expected way.

\subsection{Exponentials}

Given a game $A$, we define a game $\oc A$ as follows:
\begin{itemize}
  \item $M_{\oc A} = M_A\times\omega$.
  \item $\lambda_{\oc A} = \lambda_A \circ \pr_1$.
  \item $P_{\oc A} = \{s\in M_{\oc_A}^* \suchthat \text{$s$ is alternating},\;s\vert_n\in P_A\forall n\}$.
\end{itemize}

It can be shown \cite{ajmPcf} that this construction gives rise to a \emph{linear exponential comonad} on $\G$, and therefore that the co-Kleisli category $\G^{\oc}$ is a Cartesian closed category.

\subsection{Modelling programming languages}

The results from \cite{ajmPcf} that we will be interested are the following:
\begin{itemize}
  \item $\G_{hf}$ is a fully abstract model of PCF.
  \item $\G_{rec}$ is also a fully abstract model of PCF.  
    Moreover, it is a \emph{universal} model: every morphism between the denotation of types is the denotation of a PCF term.
\end{itemize}

We will also be interested in a result of Laird \cite{LairdCofCommCom}, which tells us about the natural language for $\G$ itself:
\begin{itemize}
  \item $\G$ is a fully abstract model of $\lambda_{\text{co}}$, an extension of PCF with coroutines.
    In particular, $\G$ is a sound and adequate model of a stateful language such as Idealized Algol.
\end{itemize}

This result -- that our model of PCF may be embedded inside a game semantics model of a stateful language -- will be useful when we come to prove Computational Adequacy for our model.  
It is a result specific to AJM games.  
For Hyland-Ong games, the category of games and \emph{innocent} strategies, which is our model of PCF, may be embedded into the category of games and \emph{visible} strategies, which is actually a fully abstract model of Idealized Algol.  
However, we will only need soundness and adequacy, so we can carry out our proof with AJM games as well.

\section{Strategies for Nondeterminism}

The results that we have quoted in the previous section all apply to \emph{deterministic} PCF.  
For nondeterministic PCF, we follow the approach of Harmer and McCusker \cite{mcCHFiniteND}, who modelled nondeterministic programs using \emph{nondeterministic strategies}; i.e., strategies where we have relaxed the nondeterminism condition.  
So, for example, if we model the type $\nat$ by the game $\bN$ with maximal positions $qn$ for $n\in\bN$, where $q$ is a fixed opening move, then we can model the term $\wn$ by the strategy $\top_\nat$ that is the whole of $P_{\bN}$.

However, this model is not yet rich enough to model may-and-must equivalence.  
Consider, for example, the following two of terms of type $\nat$:
\begin{gather*}
  \IfO \wn\;\Omega\;\0 \\
  \0\,.
\end{gather*}
In the first case, there are two possibilities: either the program returns $\0$ or it diverges.
In the second case, the program always returns $\0$.

Since divergence is represented by \emph{partiality} in the strategy, these terms both have the same denotation: the strategy with unique maximal play $q0$.  
However, they are not must-equivalent, since $\0\mustconverge$, while the other term may diverge.  
The point is that in the first case the addition of the play $q0$ into the strategy obscures the partiality present in the denotation of $\Omega$.

Harmer and McCusker's solution to this problem is to augment each strategy by the addition of a set of \emph{divergences}: $O$-positions at which we signify that play may diverge.

Tracking divergences explicitly in this way requires some care when we compose strategies.  
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))\,,
  \]
where $n;P$ is a shorthand for $\IfO n\;P\;P$, is given by a total strategy, without divergences: namely the strategy with plays of the form
\[
  \begin{array}{cc}
    \mathbf{\bN} & \mathbf{\bN} \\
    & q \\
    q & \\
    n_1 & \\
    q & \\
    n_2 & \\
    \vdots &
  \end{array}\,.
  \]
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.

@@

This works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy with maximal plays of the form:
\[
  \begin{array}{cc}
    \mathbf{\bN} & \mathbf{\bN} \\
    & q \\
    q & \\
    n_1 & \\
    \vdots & \\
    q & \\
    n_m & \\
    & 0
  \end{array}\,.
  \]
Note that this strategy strictly contains the one we considered before, and therefore that the denotation of
\[
  \IfO \wn M N
  \]
has the same denotation as $N$, even though it has all the divergent evaluations of $M$, while $N\n\mustconverge$ for all $\n$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergence into the composite -- even though no divergence occurs in the evaluation of $N$.

@@

\subsection{Strategies}

Here, we follow \cite{mcCHFiniteND}, but with the addition of infinite positions.
Given a set $X\subset M^*$, we write $\bar{X}$ for the ideal closure of $X$; i.e., $X$ together with the set of all infinite sequences $s\in M^\omega$ all of whose finite prefixes are contained in $X$.

A \emph{strategy} for a game $A=(A,\zeta_A)$ is given by a pair $(T_\sigma, D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a prefix-closed subset of $\bar{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$, then $sa\in T_\sigma$.  
    In other words, $T_\sigma$ a nondeterministic strategy in the usual sense (possibly incorporating infinite plays).  
    We consider all infinite positions in $A$ to be $O$-positions.
  \item $D_\sigma\subset T_\sigma$ is a \emph{postfix}-closed set of $O$-positions satifying the following conditions:
    \begin{description}
      \item[All divergences arise from positions] If $d$ is minimal in $D_\sigma$ then either $d\in T_\sigma$ or $d$ is an infinite position that is the limit of positions in $T_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or there exists $sa\in T_\sigma$.
      \item[Infinite positions are divergences] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

We say that a strategy $\sigma$ is \emph{complete} if $T_\sigma=\bar{T_\sigma}$; i.e., if $s\in T_\sigma$ whenever $s$ is an infinite position, all of whose finite prefixes live in $T_\sigma$.  
For example, the denotation of $M$ above is complete, but the denotation of $N$ is not complete.

We say that $\sigma$ is \emph{locally complete} if $T_\sigma$ can be written as the union of complete strategies.

It is easy to see that the denotation of any term from nondeterministic PCF must be locally complete.  
An example of a strategy that is not locally complete is given by taking the denotation of $M$ above and considering it as a strategy with no infinite positions.  
The difference between this strategy and the denotation of $N$, is that every finite position in $N$ is eventually completed in the copy of $\bN$ on the right, so it is still locally complete.

We will require our strategies to be locally complete from here on.  
The problem with allowing non-locally complete strategies is that their composition does not necessarily satisfy the diverge-or-reply rule.

\subsection{Composition of strategies}

Let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be strategies.  
We define $T_{\sigma;\tau}$ in the usual way.  
In order to define $D_{\sigma;\tau}$, we first define
\[
  \sigma\dv\tau = \pocl\left\{\s\in\seqs{(M_A\cprd M_B\cprd M_C)}\;\middle|\; \mbox{\pbox{\textwidth}{\textbf{either} $\s\vert_{A,B}\in D_\sigma$\\ \hspace*{12pt}and $s\vert_{B,C}\in T_\tau$ \\ \textbf{or} $\s\vert_{A,B}\in T_\sigma$ \\\hspace*{12pt}and $s\vert_{B,C}\in D_\tau$}}\right\}\,,
  \]
where $\pocl X$ denotes the \emph{postfix closure} of $X$; i.e., the set of all $O$-positions in $A\implies C$ that have some prefix in $X$.

Lastly, we define
\[
  D_{\sigma;\tau} = \{\s\vert_{A,C}\suchthat\s\in\sigma\dv\tau\}\,.
  \]

We need to show firstly that composition preserves local completeness.  
This is not completely trivial, since the composition of \emph{complete} strategies need not be complete: for example, we may write $N=N'\;\wn$, where $N'$ is a deterministic strategy.  
Both $N'$ and $\wn$ have complete denotations, but their composite, $N$, has a denotation that is only locally complete.

However, we can show that the composition of \emph{deterministic} complete strategies $\sigma$ and $\tau$ is complete.  
The proof makes use of a lemma from \cite{abramskyjagadeesangames} that says that if $s\in\sigma;\tau$ is some position, where $\sigma$ and $\tau$ are both deterministic, then there is a unique minimal position $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.  
This means that if $s_1\prefix s_2\prefix \cdots$ is an infinite increasing sequence in $\sigma;\tau$, then there is a corresponding infinite increasing sequence $\s_1\prefix\s_2\prefix\cdots$ in $\sigma\|\tau$, whose limit must be contained in $\sigma\|\tau$ by completeness of $\sigma$ and $\tau$.  
It follows that the limit of $s_1,s_2,\dots$ is contained in $\sigma;\tau$.

Now the fact that composition preserves local completeness is an easy consequence of monotonicity of composition with respect to the prefix ordering, together with the fact that if a strategy $\sigma$ is locally complete -- i.e., if it is the union of some collection $(\sigma_\alpha)$ of complete strategies, then the $\sigma_\alpha$ may always be chosen to be deterministic.

>>>>

\section{Idealized Algol with Countable Nondeterminism}

The language that we will be modelling is the language Idealized Algol \cite{SamsonGuyIAPassive}, extended with an additional constant $\wn$ representing countable nondeterminism.  
This is similar to the approach adopted in \cite{mcCHFiniteND}, which extends Idealized Algol with \emph{finite} nondeterminism.  
The types of the language are defined inductively as follows:
\[
  T \Coloneqq \nat \mid \com \mid \Var \mid T \to T\,.
  \]
Meanwhile, the terms are those given in \cite{SamsonGuyIAPassive}, together with the nondeterministic choice:
\begin{IEEEeqnarray*}{RCL}
  M & \Coloneqq & x \mid \lambda x . M \mid M\;M \mid \Y_T M \mid \\
  && \n \mid \skipp \mid \suc M \mid \pred M \mid \\
  && \IfO M\;M\;M \mid M;M \mid \\
  && M \coloneqq M \mid \deref M \mid \\
  && \neww_T\;M \mid \mkvar M\;M \mid \wn\,.
\end{IEEEeqnarray*}

The typing rule for $\wn$ is $\Gamma \vdash \wn\from\nat$.
We shall use the letter $v$ to range over variables of type $\Var$.

The value forms in the language are given by
\[
  V \Coloneqq \skipp \mid \n \mid \lambda x.M \mid v \mid \mkvar
  \]
We define predicates $M \converges V$ ($M$ may converge to $V$) and $M\mustconverge$ ($M$ must converge) inductively to give a big-step operational semantics for our language.  
We give a selection of the appropriate rules in Figure \ref{fig:ia-os}; this presentation is almost identical to that given in \cite{mcCHFiniteND}, with a slightly different rule for the countable nondeterminism.  
In each rule, $\ia{s}{M}$ is a \emph{configuration} of the language, where $M$ is a term, and $s$ is a \emph{store}; i.e., a function from the set of variables free in $M$ to the set of natural numbers.  
If $s$ is a store and $v$ a variable, we write $\stup s v n$ for the state formed by updating the value of the variable $v$ to $n$.

\begin{figure*}
  \begin{mathpar}
    \inferrule*{\ia s N \converges \ia {s'} \n \\ \ia {s'} M \converges \ia {s''} v}
    {\ia s {M \coloneqq N} \converges \ia {\stup {s''} v n} \skipp}
    \and
    \inferrule*[right=${s'(v) = n}$]{\ia s M \converges \ia {s'} v}
    {\ia s {\deref M} \converges \ia {s'} \n}
    \and
    \inferrule*{\ia s M \converges \ia {s'} \skipp \\ \ia {s'} N \converges \ia {s''} V}
    {\ia s {M;N} \converges \ia {s''} V}
    \and
    \inferrule*{\ia {\stup s v 0} M \converges \ia {\stup{s'} v n} V}
    {\ia s {\new_T\;\lambda v.M} \converges \ia {s'} V}
    \and
    \inferrule*[right=$n\in\bN$]{\ia s {M \n} \converges \ia {s'} V}
    {\ia s {M \wn} \converges \ia {s'} V}
    \and
    \inferrule*{\forall n\in\bN.\ia s {M \n}\mustconverge}
    {\ia s {M \wn} \mustconverge}
    \and
    \inferrule*{\ia s M \mustconverge \\ \forall \ia {s'} {\lambda x.M'}\esuchthat \ia s M \converges \ia {s'} {\lambda x.M'} \Rightarrow \ia {s'}{M'[N/x]}\mustconverge}
    {\ia s {M\;N}\mustconverge}
  \end{mathpar}
  \caption{Operational semantics of Idealized Algol with Countable Nondeterminism}
  \label{fig:ia-os}
\end{figure*}

Given a term $M$ of ground type $\com$ or $\nat$, a proof $\pi$ that $\ia s M\converges V$ gives rise to a (possibly infinite) sequence of natural numbers corresponding to a bottom-to-top, left-to-right reading of the natural numbers $n$ used in the first rule for $\wn$.  
We call such a sequence an \emph{evaluation} of the configuration $\ia s M$.  
It is perhaps easier to view this from the perspective of a small-step reduction: the sequence of numbers is formed by listing the different values that the nondeterminism constant $\wn$ decays to over the course of the evaluation $\pi$.

Let $T$ be an Idealized Algol type, and let $M,N\from T$ be terms.  
Then we write $M\lmam N$ if for all compatible contexts $C[-]$ of ground type we have
\begin{gather*}
  C[M]\converges V \Rightarrow C[N] \converges V\\
  C[M]\mustconverge \Rightarrow C[N] \mustconverge
\end{gather*}

We write $M\emam N$ if $M\lmam N$ and $N\lmam M$.

@@

\section{Game Semantics}

We will use the Hyland-Ong version of Game Semantics, as in \cite{SamsonGuyIAPassive}.

\subsection{Arenas}

An \emph{arena} is given by a triple $A=(M_A,\lambda_A,\ts_A)$, where
\begin{itemize}
  \item $M_A$ is a countable set of moves,
  \item $\lambda_A\from m_A\to\OP\times\QA$ designates each move as either an \emph{$O$-move} or a \emph{$P$-move}, and as either a \emph{question} or an \emph{answer}.  
    We define $\lambda_A^{OP}=\pr_1\circ\lambda_A$ and $\lambda_A^{QA}=\pr_2\circ\lambda_A$.  
    We also define $\neg\from\OP\times\QA\to\OP\times\QA$ to be the function that reverses the values of $O$ and $P$ while leaving $\QA$ unchanged.
  \item $\ts_A$ is an \emph{enabling relation} between $M_A+\{*\}$ and $M_A$ satisfying the following rules:
    \begin{itemize}
      \item If $a\ts_A b$ and $a\ne b$, then $\lambda_A^{OP}(a)\neq\lambda_A^{OP}(b)$.  
      \item If $*\ts_A a$, then $\lambda_A(a)=OQ$ and $b\nts_A a$ for all $b\in M_A$.
      \item If $a\ts_A b$ and $b$ is an answer, then $a$ is a question.
    \end{itemize}
    We say that a move $a\in M_A$ is \emph{initial} in $A$ if $*\ts_A a$.
\end{itemize}

Our base arenas will be the \emph{flat arenas} for the types $\nat$ and $\com$.  
Given a set $X$, the flat arena on $X$ is the arena with a single $O$-question $q$ and a $P$-answer $x$ for each $x\in X$, where $*\ts q$ and $q\ts x$ for each $x$.  
The denotation of the type $\nat$ will be the flat arena $\bN$ on the set of natural numbers, while the denotation of the type $\com$ will be the flat arena $\bC$ on the singleton set $\{a\}$.

Given an arena $A$, a \emph{justified string} in $A$ is a sequence $s$ of moves in $A$, together with \emph{justification pointers} that go from move to move in the sequence in such a way that every non-initial move $m$ in $s$ has exactly one justification pointer going back to an earlier move $n$ in $s$ such that $n\ts_A m$.  
We say that $n$ \emph{justifies} $m$.  
It is easy to see that every justified string must begin with an initial move, an hence with an $O$-question.

A \emph{legal play} $s$ is a justified string in $A$ that strictly alternates between $O$-moves and $P$-moves and is such that the corresponding $QA$-sequence formed by applying $\lambda_A^{QA}$ to terms is well-bracketed.
We write $L_A$ for the set of legal plays in $A$.

\subsection{Games and strategies}

We follow the approach taken by Abramsky and McCusker \cite{SamsonGuyIAPassive} -- a middle road between the \emph{arenas} of Hyland and Ong and the \emph{games} of \cite{ajmPcf} that makes the linear structure more apparent.

Let $s$ be a legal play in some arena $A$.  
If $m$ and $n$ are moves in $s$ such that there is a chain of justification pointers leading from $m$ back to $n$, we say that $n$ \emph{hereditarily justifies} $m$.  
Given some set $S$ of initial moves in $s$, we write $s\vert_S$ for the subsequence of $s$ made up of all those moves that are hereditarily justified by some move in $S$.

A \emph{game} is a tuple $A=(M_A,\lambda_A,\ts_A,P_A)$, where $(M_A,\lambda_A,\ts_A)$ is an arena and $P_A$ is a non-empty prefix-closed set of legal plays in that arena such that if $s\in P_A$ and $I$ is a non-empty set of initial moves in $s$, then $s\vert_I\in P_A$.

\subsubsection{Multiplicatives}

Let $A,B$ be games.  
We define games $A\tensor B$ and $A\implies B$ as follows.
\begin{IEEEeqnarray*}{RCL}
  M_{A\tensor B} & \quad=\quad & M_A + M_B\,. \\
  \lambda_{A\tensor B} & = & [\lambda_A,\lambda_B]\,.\\
  *\ts_{A\tensor B} n & \Leftrightarrow & * \ts_A n \text{ or } * \ts_B n\,. \\
  m \ts_{A\tensor B} n & \Leftrightarrow & m \ts_A n\text{ or } m \ts_B n\,.\\
  P_{A\tensor B} & = & \{s\in L_{A\tensor B}\suchthat s\vert_A\in P_A\text{ and }s\vert_B\in P_B\}\,.
\end{IEEEeqnarray*}
\begin{IEEEeqnarray*}{RCL}
  M_{A\implies B} & \quad=\quad & M_A + M_B\,. \\
  \lambda_{A\implies B} & = & [\neg\circ\lambda_A,\lambda_B]\,.\\
  *\ts_{A\implies B} n & \Leftrightarrow & * \ts_B m\,.\\[1.0ex]
  m \ts_{A\implies B} n & \Leftrightarrow & \mbox{\pbox\textwidth{$m \ts_A n$ or $m\ts_B n$ \\ or (for $m\neq *$) $ * \ts_B m$ and $* \ts_A n$\,.}} \\[1.0ex]
  P_{A\implies B} & = & \{s\in L_{A\implies B}\suchthat s\vert_A\in P_A\text{ and }s\vert_B\in P_B\}\,.
\end{IEEEeqnarray*}

\subsubsection{Modelling countable nondeterminism}

Our definition of a strategy will be modelled upon that given in \cite{mcCHFiniteND}.  
We model nondeterministic computations by relaxing the determinism constraint on strategies -- so player $P$ may have multiple replies to any given $O$-move.

In addition, we have to keep track of any possible divergence in the computation; this is so we can distinguish terms such as
\begin{mathpar}
  \IfO \wn\;\Omega\;\0 \and \0\,,
\end{mathpar}
where the term on the right must converge (to $\0$), while the term on the left has a possible divergence.  
The traditional way of representing divergences in game semantics is by a partiality in the strategy; i.e., an $O$-move to which $P$ has no reply, but this partiality will be obscured by the alternative behaviour in the denotation of the strategy on the left.

We follow \cite{mcCHFiniteND} by modelling a strategy as a pair $(T_\sigma,D_\sigma)$, where $T_\sigma$ is a nondeterministic strategy in the usual sense and $D_\sigma$ is a set of $O$-plays after which there is a possibility of divergence.

Tracking divergences explicitly in this way requires some care when we compose strategies.  
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))\,,
  \]
where $n;P$ is a shorthand for $\IfO n\;P\;P$, is given by a total strategy, without divergences: namely the strategy $\mu$ with plays of the form
\[
  \begin{array}{cc}
    \mathbf{\bN} & \mathbf{\bN} \\
    & q \\
    q & \\
    n_1 & \\
    q & \\
    n_2 & \\
    \vdots &
  \end{array}\,.
  \]
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.

@@

This works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of the form:
\[
  \begin{array}{cc}
    \mathbf{\bN} & \mathbf{\bN} \\
    & q \\
    q & \\
    n_1 & \\
    \vdots & \\
    q & \\
    n_m & \\
    & 0
  \end{array}\,.
  \]
Note that this strategy strictly contains the one we considered before, and therefore that the denotation of
\[
  \IfO \wn M N
  \]
has the same denotation as $N$, even though it has all the divergent evaluations of $M$, while $N\n\mustconverge$ for all $\n$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergence into the composite -- even though no divergence occurs in the evaluation of $N$.

@@

\subsubsection{Strategies}

Given an arena $A$, we define an \emph{infinite justified string} in the obvious way.  
We define $\bar{P_A}$ to be $P_A$ together with the set of all those infinite justified sequences that have all finite prefixes in $P_A$.

Let $A$ be a game.  
A \emph{strategy} $\sigma$ for $A$ is a pair $(T_\sigma,D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a non-empty prefix-closed subset of $\bar{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ then $sa\in T_\sigma$.
  \item $D_\sigma\subset \bar{P_A}$ is a postfix-closed set of plays in $\bar{P_A}$ that either end with an $O$-move or are infinite.  
    We require $D_\sigma$ to obey the following rules:
    \begin{description}
      \item[Divergences come from plays] If $d\in D_\sigma$ then there exists some $s\prefix d$ such that $s\in T_\sigma\cap D_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or $sa\in T_\sigma$ for some legal play $sa$.
      \item[Infinite positions are divergent] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

\subsubsection{Composition of strategies}

Given games $A,B,C$, we define a justified string over $A,B,C$ to be a sequence $\s$ of moves with justification pointers from all moves except the initial moves in $C$.  
Given such a string, we may form the restrictions $\s\vert_{A,B}$ and $\s\vert_{B,C}$ by removing all moves in either $C$ or $A$, together with all justification pointers pointing into these games.  
We define $\s\vert_{A,C}$ to be the sequence formed by removing all moves from $B$ from $\s$ and all pointers to moves in $B$, \emph{unless} we have a sequence of pointers $a \to b \to c$, in which case we replace them with a pointer $a \to c$.

We call such a sequence $\s$ a \emph{legal interaction} if $\s\vert_{A,B}\in P_{A\implies B}$, $\s\vert_{B,C}\in P_{B\implies C}$ and $\s\vert_{A,C}\in P_{A\implies C}$.  
We write $\Int_\infty(A,B,C)$ for the set of (possibly infinite) legal interactions between $A$, $B$ and $C$.

Now, given strategies $\sigma\from A \implies B$ and $\tau\from B \implies C$, we define
\[
  T_\sigma\|T_\tau = \{\s\in\Int_\infty(A,B,C)\suchthat \s\vert_{A,B}\in T_\sigma,\;\s\vert_{B,C}\in T_\tau\}\,,
  \]
and then set
\[
  T_{\sigma;\tau} = \{\s\vert_{A,C}\suchthat \s\in T_\sigma\|T_\tau\}\,.
  \]
As for divergences in $\sigma;\tau$, our approach is actually simpler than that in \cite{mcCHFiniteND}; we set
\[
  D_\sigma\dv D_\tau = \left\{\s\in \Int_\infty(A,B,C)\;\middle|\; \mbox{\pbox{\textwidth}{\textbf{either} $\s\vert_{A,B}\in D_\sigma$\\ \hspace*{12pt}and $s\vert_{B,C}\in T_\tau$ \\ \textbf{or} $\s\vert_{A,B}\in T_\sigma$ \\\hspace*{12pt}and $s\vert_{B,C}\in D_\tau$}}\right\}\,.
  \]
We then set
\[
  D_{\sigma;\tau} = \pocl_{A\implies C}\{\s\vert_{A,C}\suchthat\s\in D_\sigma\dv D_\tau\}\,,
  \]
where $\pocl X$ denotes the \emph{postfix closure} of $X$; i.e., the set of all $O$-plays in $P_{A\implies C}$ that have some prefix in $X$.

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself a legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

We need to impose one more condition on strategies:
\begin{definition}
  Let $\sigma$ be a strategy for a game $A$.  
  We say that $\sigma$ is \emph{complete} if $T_\sigma=\bar{T_\sigma}$; i.e., $T_\sigma$ contains an infinite position $s$ if it contains every finite prefix of $s$.  
\end{definition}

Any finite-nondeterminism strategy in the sense of \cite{mcCHFiniteND} may be interpreted as a complete strategy by enlarging it with all its infinite limiting plays.  
However, when we introduce countable nondeterminism, we introduce the possibility of strategies that are not complete.  
For example, the strategy $\nu$ that we mentioned above has an infinite increasing sequence of plays $q0\prefix q0q0\prefix \cdots$, but has no infinite play corresponding to its limit.  
Nonetheless, we do not want to allow arbitrary strategies: for example, the strategy $\mu$ above should include the infinite play $qq0q0\dots$; the strategy $\mu^\circ$ formed by removing this infinite play has no meaning in our language.  
Indeed, if we compose $\mu^\circ$ with the strategy $\0$ for $\bN$ on the left, then the resulting strategy does not satisfy diverge-or-reply.

\begin{definition}
  Let $\sigma$ be a strategy for a game $A$.  
  We say that $\sigma$ is \emph{locally complete} if it may be written as the union of complete strategies; i.e., there exist $\sigma_\alpha$ such that $T_\sigma=\bigcup T_{\sigma_\alpha}$ and $D_\sigma=\bigcup D_{\sigma_\alpha}$.
\end{definition}

From now on, we will use `strategy' to mean \emph{locally complete strategy}.

We need to show that the composition of locally complete strategies is locally complete.  
Note that this does not hold for \emph{complete} strategies: for example, our term $N$ above can be written as $N'\;\wn$, where $N'$ is a deterministic term with complete denotation $\nu'$.  
Then we have $\nu=\top_{\bN};\nu'$, but $\nu$ is not complete.
However, we can show that the composition of \emph{deterministic} complete strategies is complete; since a locally complete strategy may always be written as the union of complete deterministic strategies, this is sufficient to show that the composition of locally complete strategies is locally complete.
